Precise expressions for random projections: \\ Low-rank approximation and randomized Newton

### 2022-11-23 design/equivalents_nips/neurips_2020.tex
Updates from Overleaf
Working compile

% \documentclass{article}
\documentclass[thesis.tex]{subfiles}

@@

% \begin{abstract}
It is often desirable to reduce the dimensionality of a large dataset
by projecting it onto a low-dimensional subspace.  Matrix sketching
has emerged as a powerful technique for performing such dimensionality
reduction very efficiently.  Even though there is an extensive
literature on the worst-case performance of sketching, existing
guarantees are typically very different from what is observed in
practice.  We exploit recent developments in the spectral analysis of
random matrices to develop novel techniques that provide provably
accurate expressions for the expected value of random projection
matrices obtained via sketching.  These expressions can be used to
characterize the performance of dimensionality reduction in a variety
of common machine learning tasks, ranging from low-rank approximation
to iterative stochastic optimization.  Our results apply to several
popular sketching methods, including Gaussian and Rademacher sketches,
and they enable precise analysis of these methods in terms of spectral
properties of the data.  Empirical results show that the expressions
we derive reflect the practical performance of these sketching
methods, down to lower-order effects and even constant factors. 
% \end{abstract}

\section{Introduction}
Many settings in modern machine learning, optimization and scientific
computing require us to work with data matrices that are so large that
some form of dimensionality reduction is a necessary component of the
process. One of the most popular families of methods for dimensionality reduction,
coming from the literature on Randomized Numerical Linear Algebra
(RandNLA), consists of data-oblivious sketches \cite{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. 
Consider a large $m\times n$
matrix $\A$. A \emph{data-oblivious sketch} of size $k$ is the matrix $\S\A$,
where $\S$ is a $k\times m$ random matrix such that
$\E[\frac1k\S^\top\S]=\I$, whose distribution does not
depend on $\A$. This sketch reduces the first dimension of $\A$ from
$m$ to a much smaller $k$ (we assume without loss of generality that
$k \ll n \le m$), and an analogous procedure can be defined for
reducing the second 
dimension as well. This approximate representation of $\A$ is central
to many algorithms in areas such as linear regression, low-rank approximation, kernel methods,
and iterative second-order optimization. While there is a long line of research
aimed at bounding the worst-case approximation error of such
representations, these bounds are often too loose to reflect accurately
the practical performance of these methods. In this paper, we develop new theory which
enables more precise analysis of the accuracy of sketched data
representations.

A common way to measure the accuracy of the sketch $\S\A$ is by
considering the $k$-dimensional subspace spanned by its rows. The goal
of the sketch is to choose a subspace that best aligns with the
distribution of all of the $m$ rows of $\A$ in $\R^n$. Intuitively, our goal
is to minimize the (norm of the) residual when projecting a vector $\a\in\R^n$ onto that
subspace, i.e., $\a - \P\a=(\I-\P)\a$, where
$\P = (\S\A)^\dagger\S\A$ is the orthogonal projection matrix onto the subspace
spanned by the rows of $\S\A$ (and $(\cdot)^\dagger$ denotes the
Moore-Penrose pseudoinverse). For this reason, the quantity that has appeared 
ubiquitously in the error analysis of RandNLA sketching is what we call the residual
projection matrix:
\begin{align*}
  \textbf{(residual projection matrix)}\quad \P_{\!\perp}\ :=\ \I - \P\ =\ \I
  - (\S\A)^\dagger\S\A.
\end{align*}
Since $\P_{\!\perp}$ is random, the average performance of the sketch
can often be characterized by its expectation, $\E[\P_{\!\perp}]$.
For example, the low-rank approximation error of the
sketch can be expressed as
$\E[\|\A - \A\P\|_F^2]= \tr\,\A^\top\A\,\E[\P_{\!\perp}]$, where
$\|\cdot\|_F$ denotes the Frobenius norm.  A similar formula follows for
the trace norm error of a sketched Nystr\"om approximation
\cite{Williams01Nystrom,revisiting-nystrom}.
Among others, this approximation error
appears in the analysis of sketched kernel 
ridge regression \cite{fanuel2020diversity} and Gaussian process
regression \cite{sparse-variational-gp}. Furthermore, a variety of
iterative algorithms, such as randomized second-order methods for
convex optimization \cite{Qu2015Feb,Qu2016,Gower2019,jacsketch}
and linear system solvers based on 
the generalized Kaczmarz method \cite{generalized-kaczmarz},
have convergence guarantees which depend on the extreme eigenvalues of
$\E[\P_{\!\perp}]$. Finally, a generalized form of the expected
residual projection has been recently used to model the implicit
regularization of the interpolating solutions in over-parameterized
linear models \cite{surrogate-design,BLLT19_TR}.

\subsection{Main result}
Despite its prevalence in the literature, the 
expected residual projection is not well understood, even in such
simple cases as when $\S$ is a Gaussian sketch (i.e., with
i.i.d.~standard normal entries). We address this by providing a
surrogate expression, i.e., a simple analytically
tractable approximation, for this matrix quantity:
\begin{align}
  \E[\P_{\!\perp}]\ \overset\epsilon\simeq \ \bar\P_{\!\perp}:=(\gamma\A^\top\A +
  \I)^{-1},\quad\text{with \ $\gamma>0$ \ s.t. \ }\tr\,\bar\P_{\!\perp} = n-k.\label{eq:surrogate}
\end{align}
Here, $\overset{\epsilon}{\simeq}$ means that while the surrogate expression is
not exact, it approximates the true quantity up to some $\epsilon$
accuracy. Our main result provides a rigorous approximation guarantee
for this surrogate expression with respect to a
range of sketching matrices $\S$, including the standard Gaussian and
Rademacher sketches. We state the result using the
positive semi-definite ordering denoted by $\preceq$. 
\begin{theorem}\label{t:main}
Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
  we let $\rho = r/k$ be a fixed constant larger than $1$, then
  \begin{align*}
    (1-\epsilon)\,\bar\P_{\!\perp}\preceq\E[\P_{\!\perp}]\preceq
    (1+\epsilon)\,\bar\P_{\!\perp}\quad\text{for}\quad \epsilon =
    O(\tfrac1{\sqrt r}).
  \end{align*}
\end{theorem}
In other words, when the sketch size $k$ is smaller than the stable rank $r$
of $\A$, then the discrepancy between our surrogate expression
$\bar\P_{\!\perp}$ and $\E[\P_{\!\perp}]$ is of the order
$1/\sqrt r$, where the big-O notation hides only the dependence
on $\rho$ and on the sub-gaussian constant (see Theorem \ref{t:main-tech} for
more details). Our proof of Theorem \ref{t:main} is inspired by the techniques from
random matrix theory which have been used to analyze the asymptotic
spectral distribution of large random matrices by focusing on the
associated matrix resolvents and Stieltjes transforms
\cite{hachem2007deterministic,bai2010spectral}. However, our analysis
is novel in several respects:
\begin{enumerate}
  \item The residual projection matrix can be obtained
    from the appropriately scaled resolvent matrix $z(\A^\top\S^\top\S\A+z\I)^{-1}$ by
    taking $z\rightarrow 0$. Prior work (e.g.,
      \cite{HMRT19_TR}) combined this with an exchange-of-limits argument to
      analyze the asymptotic behavior of the residual projection. 
      This approach, however, does not allow for a precise control in
      finite-dimensional problems. We are able to provide a more
      fine-grained, non-asymptotic analysis by working directly with
      the residual projection itself, instead of the resolvent.
  \item We require no assumptions on the largest and smallest singular
    value of $\A$. Instead, we derive our bounds in terms of the
    stable rank of $\A$ (as opposed to its actual rank), which
    implicitly compensates for ill-conditioned data matrices.
  \item We obtain upper/lower bounds for $\E[\P_{\!\perp}]$ in terms
    of the positive semi-definite ordering~$\preceq$, which can be
    directly converted to guarantees for the precise
    expressions of expected low-rank approximation error derived in
    the following section. 
    % are stronger than the standard error bounds
    % $\|\E[\P_{\!\perp}]-\bar\P_{\!\perp}\|$ because they lead to relative error
    % approximations for \emph{all} of the eigenvalues of
    % $\E[\P_{\!\perp}]$ (with respect to those of $\bar \P_{\!\perp}$).
  \end{enumerate}

It is worth mentioning that the proposed analysis is significantly
different from the sketching literature based on subspace
embeddings (e.g.,
\cite{sarlos-sketching,cw-sparse,nn-sparse,projection-cost-preserving,optimal-matrix-product}),
in the sense that here our object of interest is not to obtain
a worst-case approximation with high probability, but rather, our
analysis provides \emph{precise} characterization on the
\emph{expected} residual projection matrix that goes \emph{beyond
  worst-case bounds}. From an application perspective, the subspace
embedding property is neither sufficient nor necessary for many
numerical implementations of sketching \citep{blendenpik,lsrn}, 
or statistical results \citep{GarveshMahoney_JMLR,dobriban2019asymptotics,yang2020reduce}, as well as in the context of iterative optimization and implicit regularization (see Sections~\ref{s:newton}~and~\ref{s:implicit} below), which are discussed in detail as concrete applications of the proposed analysis.

\subsection{Low-rank approximation}
\label{s:low-rank}
We next provide some immediate corollaries of Theorem~\ref{t:main},
where we use $x\overset\epsilon\simeq y$ to denote a 
multiplicative approximation $|x-y|\leq \epsilon y$.
Note that our analysis is new even for the classical Gaussian sketch where
the entries of $\S$ are i.i.d.~standard normal. However the results
apply more broadly, including a standard class of data-base friendly
Rademacher sketches where each entry $s_{ij}$ is a $\pm1$ Rademacher
random variable \cite{achlioptas2003database}. We start by analyzing the Frobenius
norm error $\|\A-\A\P\|_F^2=\tr\,\A^\top\A\,\P_{\!\perp}$ of
sketched low-rank approximations. Note that by the definition of $\gamma$ in
\eqref{eq:surrogate}, we have $k =
\tr\,(\I-\bar\P_{\!\perp})=\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$,
so the surrogate expression we obtain for the expected error is
remarkably simple.
\begin{corollary} %[low-rank approximation]
  \label{c:low-rank}
Let $\sigma_i$ be the singular values of $\A$. Under the assumptions of Theorem \ref{t:main}, we have:\vspace{-1mm}
  \begin{align*}
    \E\big[\|\A-\A\P\|_F^2\big] \ \overset\epsilon\simeq\ 
    k/\gamma
\quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1} = k.
%\quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1} = k.
    % \tr\A^\top\A(\gamma\A^\top\A+\I)^{-1},
   % \quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,(\gamma\A^\top\A+\I)^{-1} = n-k.
  \end{align*}
\end{corollary}\vspace{-4mm}
\begin{remark}
  The parameter $\gamma=\gamma(k)$ increases at least linearly
  as a function of $k$, which is why the expected error will always
  decrease with increasing $k$. For example, when the singular values
  of $\A$ exhibit exponential decay, i.e., $\sigma_i^2=C\cdot\alpha^{i-1}$ for $\alpha\in(0,1)$, then the error also decreases
  exponentially, at the rate of $k/(\alpha^{-k}-1)$.
We discuss this further in Section \ref{s:explicit}, giving explicit
formulas for the error as a function of $k$ under both exponential and polynomial spectral
decay profiles.
\end{remark}
The above result is important for many RandNLA methods, and it is also relevant in the context of kernel methods,
where the data is represented via a positive semi-definite $m\times m$
kernel matrix $\K$ which corresponds to the matrix of dot-products of
the data vectors in some reproducible kernel Hilbert space. In this
context, sketching can be applied directly to the matrix $\K$ via an
extended variant of the
Nystr\"om method \cite{revisiting-nystrom}. A Nystr\"om approximation
constructed from a sketching matrix $\S$ is defined as $\tilde\K =
\C^\top\W^\dagger\C$, where $\C=\S\K$ and $\W=\S\K\S^\top$, and it is
applicable to a variety of settings, including Gaussian Process
regression, kernel machines and Independent
Component Analysis \cite{sparse-variational-gp,Williams01Nystrom,Bach2003}. By setting
$\A=\K^{\frac12}$, it is easy to see \cite{nystrom-multiple-descent} that
the trace norm error $\|\K-\tilde\K\|_*$ is identical to the squared
Frobenius norm error of the low-rank 
sketch $\S\A$, so Corollary \ref{c:low-rank} implies that
\begin{align}
  \E\big[\|\K-\tilde\K\|_*\big] \ \overset\epsilon\simeq\  k/\gamma \quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\lambda_i}{\gamma\lambda_i+1} = k,\label{eq:nystrom}
\end{align}
with any sub-gaussian sketch, where $\lambda_i$ denote the
eigenvalues of $\K$. Our error analysis 
given in Section \ref{s:explicit} is particularly relevant here, since
commonly used kernels such as the Radial 
Basis Function (RBF) or the Mat\'ern kernel induce a well-understood eigenvalue
decay \cite{Santa97Gaussianregression,RasmussenWilliams06}.

Metrics other than the aforementioned Frobenius norm error,
such as the spectral norm error \cite{tropp2011structure}, are also of significant
interest in the low-rank approximation literature. We leave these
directions for future investigation.

\subsection{Randomized iterative optimization}
\label{s:newton}
We next turn to a class of iterative methods which take
advantage of sketching to reduce the per iteration cost of
optimization. These methods have been developed in a variety of
settings, from solving linear systems to convex optimization and
empirical risk minimization, and in many cases
the residual projection matrix appears as a black box quantity whose spectral
properties determine the convergence behavior of the algorithms
\cite{generalized-kaczmarz}. With
our new results, we can precisely characterize not only the rate of
convergence, but also, in some cases, the complete evolution of
the parameter vector, for the following algorithms:
% The following algorithms
% benefit from our improved convergence analysis:
\begin{enumerate}
  \item \emph{Generalized Kaczmarz method}
    \cite{generalized-kaczmarz} for approximately solving a linear system $\A\x=\b$;
  \item \emph{Randomized Subspace Newton} \cite{Gower2019}, a second order
    method, where we sketch the Hessian matrix.
  \item \emph{Jacobian Sketching} \cite{jacsketch}, a
    class of first order methods which use additional information via a
    weight matrix $\W$ that is sketched at every iteration.
    % \item \emph{Iterative Adaptive Sketching} \cite{lacotte2019high},
    %   a second order method where the sketching matrices are
    %   adaptively rescaled to improve the convergence rate.
      % \footnote{While
      %   \cite{lacotte2019high} also use the residual projection in
      %   their analysis, adapting our techniques is less
      %   straightforward in this case.}
\end{enumerate}
 We believe that
extensions of our techniques will apply to other algorithms,
such as that of \cite{lacotte2019high}.

We next 
%Next, we 
give a result in the context of linear systems
for the generalized Kaczmarz method \cite{generalized-kaczmarz}, but a similar
convergence analysis is given for the methods of
\cite{Gower2019,jacsketch} in Appendix~\ref{a:newton}.
\begin{corollary}\label{c:kaczmarz}
Let $\x^*$ be the unique solution of $\A\x^*=\b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S\A\x=\S\b.
  \end{align*}
Under the assumptions of Theorem \ref{t:main}, with $\gamma$ defined in
\eqref{eq:surrogate} and $r=\|\A\|_F^2/\|\A||^2$, we have:
  \begin{align*}
    \E\big[\x^{t+1}-\x^*\big] \overset\epsilon\simeq 
    (\gamma\A^\top\A+\I)^{-1}\,\E\big[\x^t-\x^*\big]
    \quad\text{for}\quad\epsilon=O(\tfrac1{\sqrt r}). 
  \end{align*}
\end{corollary}
The corollary follows from Theorem \ref{t:main} combined with Theorem
4.1 in \cite{generalized-kaczmarz}. 
Note that when $\A^\top\A$ is positive definite then
$(\gamma\A^\top\A+\I)^{-1}\prec\I$, so the algorithm will
converge from any starting point, and the worst-case convergence rate of the above method can be
obtained by evaluating the largest eigenvalue of
$(\gamma\A^\top\A+\I)^{-1}$. However the result itself is much
stronger, in that it can be used to describe the (expected) trajectory of the
iterates for any starting point $\x^0$. Moreover, when the spectral decay
profile of $\A$ is known, then the explicit expressions for $\gamma$
as a function of $k$ derived in Section \ref{s:explicit} can be used
to characterize the convergence properties of generalized Kaczmarz as
well as other methods discussed above.

@@

From Lemma~\ref{l:rank-one-update}, $\P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{ \x_i^\top (\I - \P_{-i}) \x_i}$ is positive semi-definite, we have $\tr \Sigmab (\P - \P_{-i}) \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})$ that, for $L > 1$
\[
  \E |X|^L \le C_L k^{L/2}.
\]
In particular, for $L=2$, we obtain $\E |X|^2 \le C k$.
%for $c = (36 \sqrt{2})^2$ (details in proof of Theorem~3.2 in \cite{burkholder1973distribution}).

>>>>

% \documentclass{article}
\documentclass[../../thesis.tex]{subfiles}

@@

% \begin{abstract}

\chapter{Exact expectation expressions for sub-Gaussian random projections}
\label{ch:projections}

It is often desirable to reduce the dimensionality of a large dataset
by projecting it onto a low-dimensional subspace.  Matrix sketching
has emerged as a powerful technique for performing such dimensionality
reduction very efficiently. Even though there is an extensive
literature on the worst-case performance of sketching, existing
guarantees are typically very different from what is observed in
practice. Building on the Stieltjes transform methods employed
in \cref{sec:proof-of-c-projection} while establishing asymptotic
consistency of surrogate design's MSE, this chapter develops novel
techniques
that provide provably accurate expressions for the expected value of
random projection matrices obtained via sketching.  These expressions
can be used to characterize the performance of dimensionality reduction
in a variety
of common machine learning tasks, ranging from low-rank approximation
to iterative stochastic optimization.  Our results apply to several
popular sketching methods, including Gaussian and Rademacher sketches,
and they enable precise analysis of these methods in terms of spectral
properties of the data. Empirical results show that the expressions
we derive reflect the practical performance of these sketching
methods, down to lower-order effects and even constant factors.
Some of the results here were originally published in
\fullcite{precise-expressions}.

% \end{abstract}

\section{Introduction}
Many settings in modern machine learning, optimization and scientific
computing require us to work with data matrices that are so large that
some form of dimensionality reduction is a necessary component of the
process. One of the most popular families of methods for dimensionality reduction,
coming from the literature on Randomized Numerical Linear Algebra
(RandNLA), consists of data-oblivious sketches \citep{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}.
Consider a large $m\times n$
matrix $\A$. A \emph{data-oblivious sketch} of size $k$ is the matrix $\S\A$,
where $\S$ is a $k\times m$ random matrix such that
$\E[\frac1k\S^\top\S]=\I$, whose distribution does not
depend on $\A$. This sketch reduces the first dimension of $\A$ from
$m$ to a much smaller $k$ (we assume without loss of generality that
$k \ll n \le m$), and an analogous procedure can be defined for
reducing the second
dimension as well. This approximate representation of $\A$ is central
to many algorithms in areas such as linear regression, low-rank approximation, kernel methods,
and iterative second-order optimization. While there is a long line of research
aimed at bounding the worst-case approximation error of such
representations, these bounds are often too loose to reflect accurately
the practical performance of these methods. In this paper, we develop new theory which
enables more precise analysis of the accuracy of sketched data
representations.

A common way to measure the accuracy of the sketch $\S\A$ is by
considering the $k$-dimensional subspace spanned by its rows. The goal
of the sketch is to choose a subspace that best aligns with the
distribution of all of the $m$ rows of $\A$ in $\R^n$. Intuitively, our goal
is to minimize the (norm of the) residual when projecting a vector $\a\in\R^n$ onto that
subspace, i.e., $\a - \P\a=(\I-\P)\a$, where
$\P = (\S\A)^\dagger\S\A$ is the orthogonal projection matrix onto the subspace
spanned by the rows of $\S\A$ (and $(\cdot)^\dagger$ denotes the
Moore-Penrose pseudoinverse). For this reason, the quantity that has appeared
ubiquitously in the error analysis of RandNLA sketching is what we call the residual
projection matrix:
\begin{align*}
  \textbf{(residual projection matrix)}\quad \P_{\!\perp}\ :=\ \I - \P\ =\ \I
  - (\S\A)^\dagger\S\A.
\end{align*}
Since $\P_{\!\perp}$ is random, the average performance of the sketch
can often be characterized by its expectation, $\E[\P_{\!\perp}]$.
For example, the low-rank approximation error of the
sketch can be expressed as
$\E[\|\A - \A\P\|_F^2]= \tr\,\A^\top\A\,\E[\P_{\!\perp}]$, where
$\|\cdot\|_F$ denotes the Frobenius norm.  A similar formula follows for
the trace norm error of a sketched Nystr\"om approximation
\citep{Williams01Nystrom,revisiting-nystrom}.
Among others, this approximation error
appears in the analysis of sketched kernel
ridge regression \citep{fanuel2020diversity} and Gaussian process
regression \citep{sparse-variational-gp}. Furthermore, a variety of
iterative algorithms, such as randomized second-order methods for
convex optimization \citep{Qu2015Feb,Qu2016,Gower2019,jacsketch}
and linear system solvers based on
the generalized Kaczmarz method \citep{generalized-kaczmarz},
have convergence guarantees which depend on the extreme eigenvalues of
$\E[\P_{\!\perp}]$. Finally, a generalized form of the expected
residual projection has been recently used to model the implicit
regularization of the interpolating solutions in over-parameterized
linear models \citep{surrogate-design,BLLT19_TR}.

\subsection{Main result}
Despite its prevalence in the literature, the
expected residual projection is not well understood, even in such
simple cases as when $\S$ is a Gaussian sketch (i.e., with
i.i.d.~standard normal entries). We address this by providing a
surrogate expression, i.e., a simple analytically
tractable approximation, for this matrix quantity:
\begin{align}
  \E[\P_{\!\perp}]\ \overset\epsilon\simeq \ \bar\P_{\!\perp}:=(\gamma\A^\top\A +
  \I)^{-1},\quad\text{with \ $\gamma>0$ \ s.t. \ }\tr\,\bar\P_{\!\perp} = n-k.\label{eq:surrogate}
\end{align}
Here, $\overset{\epsilon}{\simeq}$ means that while the surrogate expression is
not exact, it approximates the true quantity up to some $\epsilon$
accuracy. Our main result provides a rigorous approximation guarantee
for this surrogate expression with respect to a
range of sketching matrices $\S$, including the standard Gaussian and
Rademacher sketches. We state the result using the
positive semi-definite ordering denoted by $\preceq$.
\begin{theorem}\label{t:main}
  Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
  we let $\rho = r/k$ be a fixed constant larger than $1$, then
  \begin{align*}
    (1-\epsilon)\,\bar\P_{\!\perp}\preceq\E[\P_{\!\perp}]\preceq
    (1+\epsilon)\,\bar\P_{\!\perp}\quad\text{for}\quad \epsilon =
    O(\tfrac1{\sqrt r}).
  \end{align*}
\end{theorem}
In other words, when the sketch size $k$ is smaller than the stable rank $r$
of $\A$, then the discrepancy between our surrogate expression
$\bar\P_{\!\perp}$ and $\E[\P_{\!\perp}]$ is of the order
$1/\sqrt r$, where the big-O notation hides only the dependence
on $\rho$ and on the sub-gaussian constant (see Theorem \ref{t:main-tech} for
more details). Our proof of Theorem \ref{t:main} is inspired by the techniques from
random matrix theory which have been used to analyze the asymptotic
spectral distribution of large random matrices by focusing on the
associated matrix resolvents and Stieltjes transforms
\citep{hachem2007deterministic,bai2010spectral}. However, our analysis
is novel in several respects:
\begin{enumerate}
  \item The residual projection matrix can be obtained
        from the appropriately scaled resolvent matrix $z(\A^\top\S^\top\S\A+z\I)^{-1}$ by
        taking $z\rightarrow 0$. Prior work (e.g.,
        \citep{HMRT19_TR}) combined this with an exchange-of-limits argument to
        analyze the asymptotic behavior of the residual projection.
        This approach, however, does not allow for a precise control in
        finite-dimensional problems. We are able to provide a more
        fine-grained, non-asymptotic analysis by working directly with
        the residual projection itself, instead of the resolvent.
  \item We require no assumptions on the largest and smallest singular
        value of $\A$. Instead, we derive our bounds in terms of the
        stable rank of $\A$ (as opposed to its actual rank), which
        implicitly compensates for ill-conditioned data matrices.
  \item We obtain upper/lower bounds for $\E[\P_{\!\perp}]$ in terms
        of the positive semi-definite ordering~$\preceq$, which can be
        directly converted to guarantees for the precise
        expressions of expected low-rank approximation error derived in
        the following section.
        % are stronger than the standard error bounds
        % $\|\E[\P_{\!\perp}]-\bar\P_{\!\perp}\|$ because they lead to relative error
        % approximations for \emph{all} of the eigenvalues of
        % $\E[\P_{\!\perp}]$ (with respect to those of $\bar \P_{\!\perp}$).
\end{enumerate}

It is worth mentioning that the proposed analysis is significantly
different from the sketching literature based on subspace
embeddings (e.g.,
\citep{sarlos-sketching,cw-sparse,nn-sparse,projection-cost-preserving,optimal-matrix-product}),
in the sense that here our object of interest is not to obtain
a worst-case approximation with high probability, but rather, our
analysis provides \emph{precise} characterization on the
\emph{expected} residual projection matrix that goes \emph{beyond
  worst-case bounds}. From an application perspective, the subspace
embedding property is neither sufficient nor necessary for many
numerical implementations of sketching \citep{blendenpik,lsrn},
or statistical results \citep{GarveshMahoney_JMLR,dobriban2019asymptotics,yang2020reduce}, as well as in the context of iterative optimization and implicit regularization (see Sections~\ref{s:newton}~and~\ref{s:implicit} below), which are discussed in detail as concrete applications of the proposed analysis.

\subsection{Low-rank approximation}
\label{s:low-rank}
We next provide some immediate corollaries of Theorem~\ref{t:main},
where we use $x\overset\epsilon\simeq y$ to denote a
multiplicative approximation $|x-y|\leq \epsilon y$.
Note that our analysis is new even for the classical Gaussian sketch where
the entries of $\S$ are i.i.d.~standard normal. However the results
apply more broadly, including a standard class of data-base friendly
Rademacher sketches where each entry $s_{ij}$ is a $\pm1$ Rademacher
random variable \citep{achlioptas2003database}. We start by analyzing the Frobenius
norm error $\|\A-\A\P\|_F^2=\tr\,\A^\top\A\,\P_{\!\perp}$ of
sketched low-rank approximations. Note that by the definition of $\gamma$ in
\eqref{eq:surrogate}, we have $k =
  \tr\,(\I-\bar\P_{\!\perp})=\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$,
so the surrogate expression we obtain for the expected error is
remarkably simple.
\begin{corollary} %[low-rank approximation]
  \label{c:low-rank}
  Let $\sigma_i$ be the singular values of $\A$. Under the assumptions of Theorem \ref{t:main}, we have:\vspace{-1mm}
  \begin{align*}
    \E\big[\|\A-\A\P\|_F^2\big] \ \overset\epsilon\simeq\
    k/\gamma
    \quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1} = k.
    %\quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1} = k.
    % \tr\A^\top\A(\gamma\A^\top\A+\I)^{-1},
    % \quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,(\gamma\A^\top\A+\I)^{-1} = n-k.
  \end{align*}
\end{corollary}\vspace{-4mm}
\begin{remark}
  The parameter $\gamma=\gamma(k)$ increases at least linearly
  as a function of $k$, which is why the expected error will always
  decrease with increasing $k$. For example, when the singular values
  of $\A$ exhibit exponential decay, i.e., $\sigma_i^2=C\cdot\alpha^{i-1}$ for $\alpha\in(0,1)$, then the error also decreases
  exponentially, at the rate of $k/(\alpha^{-k}-1)$.
  We discuss this further in Section \ref{s:explicit}, giving explicit
  formulas for the error as a function of $k$ under both exponential and polynomial spectral
  decay profiles.
\end{remark}
The above result is important for many RandNLA methods, and it is also relevant in the context of kernel methods,
where the data is represented via a positive semi-definite $m\times m$
kernel matrix $\K$ which corresponds to the matrix of dot-products of
the data vectors in some reproducible kernel Hilbert space. In this
context, sketching can be applied directly to the matrix $\K$ via an
extended variant of the
Nystr\"om method \citep{revisiting-nystrom}. A Nystr\"om approximation
constructed from a sketching matrix $\S$ is defined as $\tilde\K =
  \C^\top\W^\dagger\C$, where $\C=\S\K$ and $\W=\S\K\S^\top$, and it is
applicable to a variety of settings, including Gaussian Process
regression, kernel machines and Independent
Component Analysis \citep{sparse-variational-gp,Williams01Nystrom,Bach2003}. By setting
$\A=\K^{\frac12}$, it is easy to see \citep{nystrom-multiple-descent} that
the trace norm error $\|\K-\tilde\K\|_*$ is identical to the squared
Frobenius norm error of the low-rank
sketch $\S\A$, so Corollary \ref{c:low-rank} implies that
\begin{align}
  \E\big[\|\K-\tilde\K\|_*\big] \ \overset\epsilon\simeq\  k/\gamma \quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\lambda_i}{\gamma\lambda_i+1} = k,\label{eq:nystrom}
\end{align}
with any sub-gaussian sketch, where $\lambda_i$ denote the
eigenvalues of $\K$. Our error analysis
given in Section \ref{s:explicit} is particularly relevant here, since
commonly used kernels such as the Radial
Basis Function (RBF) or the Mat\'ern kernel induce a well-understood eigenvalue
decay \citep{Santa97gaussianregression,RasmussenWilliams06}.

Metrics other than the aforementioned Frobenius norm error,
such as the spectral norm error \citep{tropp2011structure}, are also of significant
interest in the low-rank approximation literature. We leave these
directions for future investigation.

\subsection{Randomized iterative optimization}
\label{s:newton}
We next turn to a class of iterative methods which take
advantage of sketching to reduce the per iteration cost of
optimization. These methods have been developed in a variety of
settings, from solving linear systems to convex optimization and
empirical risk minimization, and in many cases
the residual projection matrix appears as a black box quantity whose spectral
properties determine the convergence behavior of the algorithms
\citep{generalized-kaczmarz}. With
our new results, we can precisely characterize not only the rate of
convergence, but also, in some cases, the complete evolution of
the parameter vector, for the following algorithms:
% The following algorithms
% benefit from our improved convergence analysis:
\begin{enumerate}
  \item \emph{Generalized Kaczmarz method}
        \citep{generalized-kaczmarz} for approximately solving a linear system $\A\x=\b$;
  \item \emph{Randomized Subspace Newton} \citep{Gower2019}, a second order
        method, where we sketch the Hessian matrix.
  \item \emph{Jacobian Sketching} \citep{jacsketch}, a
        class of first order methods which use additional information via a
        weight matrix $\W$ that is sketched at every iteration.
        % \item \emph{Iterative Adaptive Sketching} \citep{lacotte2019high},
        %   a second order method where the sketching matrices are
        %   adaptively rescaled to improve the convergence rate.
        % \footnote{While
        %   \citep{lacotte2019high} also use the residual projection in
        %   their analysis, adapting our techniques is less
        %   straightforward in this case.}
\end{enumerate}
We believe that
extensions of our techniques will apply to other algorithms,
such as that of \citep{lacotte2019high}.

We next
%Next, we 
give a result in the context of linear systems
for the generalized Kaczmarz method \citep{generalized-kaczmarz}, but a similar
convergence analysis is given for the methods of
\citep{Gower2019,jacsketch} in \Cref{a:newton}.
\begin{corollary}\label{c:kaczmarz}
  Let $\x^*$ be the unique solution of $\A\x^*=\b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S\A\x=\S\b.
  \end{align*}
  Under the assumptions of Theorem \ref{t:main}, with $\gamma$ defined in
  \eqref{eq:surrogate} and $r=\|\A\|_F^2/\|\A||^2$, we have:
  \begin{align*}
    \E\big[\x^{t+1}-\x^*\big] \overset\epsilon\simeq
    (\gamma\A^\top\A+\I)^{-1}\,\E\big[\x^t-\x^*\big]
    \quad\text{for}\quad\epsilon=O(\tfrac1{\sqrt r}).
  \end{align*}
\end{corollary}
The corollary follows from Theorem \ref{t:main} combined with Theorem
4.1 in \citep{generalized-kaczmarz}.
Note that when $\A^\top\A$ is positive definite then
$(\gamma\A^\top\A+\I)^{-1}\prec\I$, so the algorithm will
converge from any starting point, and the worst-case convergence rate of the above method can be
obtained by evaluating the largest eigenvalue of
$(\gamma\A^\top\A+\I)^{-1}$. However the result itself is much
stronger, in that it can be used to describe the (expected) trajectory of the
iterates for any starting point $\x^0$. Moreover, when the spectral decay
profile of $\A$ is known, then the explicit expressions for $\gamma$
as a function of $k$ derived in Section \ref{s:explicit} can be used
to characterize the convergence properties of generalized Kaczmarz as
well as other methods discussed above.

@@

  From Lemma~\ref{l:rank-one-update}, $\P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{ \x_i^\top (\I - \P_{-i}) \x_i}$ is positive semi-definite, we have $\tr \Sigmab (\P - \P_{-i}) \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})$ that, for $L > 1$
  \[
    \E |X|^L \le C_L k^{L/2}.
  \]
  In particular, for $L=2$, we obtain $\E |X|^2 \le C k$.
  %for $c = (36 \sqrt{2})^2$ (details in proof of Theorem~3.2 in \citep{burkholder1973distribution}).
